From physical principles to ???

### 2016-04-21 2015CompleteDerivation/completeDerivation.tex
More work on classical Hamiltonian

Given that scientific reductionism (i.e. the idea of reducing physical systems and interactions to the sum of their constituent parts in order to make them easier to study) is at the heart of fundamental physics, we now explore how to characterize a system in terms of its components. That is, we want to study the relationship between the state space $\mathcal{C}$ of a composite system and the state space $\mathcal{S}$ of its parts. In general, this is quite a complicated thing to do, that requires intimate knowledge of the system at hand. So we simplify our problem and study the system is made of infinitesimal homogeneous parts. What we'll find is that under the additional assumption that the system is infinitesimally reducible (i.e. its state is equivalent to the state of its parts) and that each part undergoes deterministic and reversible evolution, the motion is suitably described by the standard framework of classical Hamiltonian mechanics.

\subsection{Homogeneous decomposable systems}
The notion that a system is decomposable means the states is equipped with a rule of composition that allows to write $\mathcal{c}=\mathcal{c}_1+\mathcal{c}_2$: the composite system is the sum of its part. For example, the state of a ball is equal to the state of its left and right parts.

The notion that the system is homogeneous means that the states of the composite and of the parts are not unrelated: they are made of the same material.\footnote{Whether a system is homogeneous depends on context (e.g. air can be thought as homogeneous if the mixture of gases does not change in space or in time due to phase transitions or chemical processes) and one must check that such property is maintained by time evolution.} In fact, the state space $\mathcal{C}$ of all systems composed of such material will include the state of the system as well as the states of all its part (i.e $\mathcal{c}, \mathcal{c}_1, \mathcal{c}_2 \in \mathcal{C}$). Since combining any two systems made of a homogeneous material will always give us a system made of the same homogeneous material (e.g. combining elements made of water gives us another element made of water), the state space $\mathcal{C}$ is closed under composition.

@@

\begin{justification}
We claim $\mathcal{C}$ is an additive monoid. There exist a law of composition $+ : \mathcal{C} \times \mathcal{C} \rightarrow \mathcal{C}$ that takes two states and returns one that is the physical composition of the two. The domain and codomain match because the material is homogeneous. The law is commutative $\mathcal{c}_1 +\mathcal{c}_2 = \mathcal{c}_2+\mathcal{c}_1$ and associative $(\mathcal{c}_1 + \mathcal{c}_2) + \mathcal{c}_3 = \mathcal{c}_1 + (\mathcal{c}_2 + \mathcal{c}_3)$, as it does not matter in what order we physically compose the parts. There exist a unique zero element $\mathcal{c} + 0 = \mathcal{c}$ and it represent the physically empty state (i.e. no amount of material).

We claim $\mathcal{C}$ is an additive group. As we want to describe changes during the evolution (i.e. $\hat{\mathcal{c}} = \mathcal{c} +\delta \mathcal{c}$) we introduce an inverse $- : \mathcal{C} \rightarrow \mathcal{C}$ such that $\mathcal{c} + ( - \mathcal{c}) = 0 \forall \mathcal{c} \in \mathcal{C}$. Such inverse may introduce objects that do not properly represent a physical state, but a state change. A change of a physically distinguishable object is itself physically distinguishable therefore $\mathcal{C}$ is still a Hausdorff and second countable topological manifold as per \ref{statedef}. We'll still call $\mathcal{C}$ state space committing an abuse of terminology.
\end{justification}

@@

We will call a classical system one that is infinitesimally reducible. That is: it is made of arbitrarily small parts, which we call particles, and the state of the whole system coincides with the sum of the states of the parts. Given the state space $\mathcal{S}$ for the infinitesimal parts, a composite state $\mathcal{c}$ will tell us the amount of material for each possible particle state. That is, each state is fully identified by a function $\rho: \mathcal{S} \to \mathbbm{R}$ that returns the amount of material in the composite state $\mathcal{c}$ that is prepared in the particle state $\mathcal{s}$. For example, given a certain configuration of gas, we can tell the amount of material that is at a specific point in space with a specific value of momentum.

The notion that the parts are infinitesimal requires the co-domain of $\rho$ to be continuous, as opposed to discrete. The state space $\mathcal{S}$, instead, can be finite. For example, for a system composed of different tanks connected by pipes, the state of the composite system could be the overall distribution of water among the tanks. The amount in each tank is continuous (as we assume the water to be infinitesimally divisible) yet there are only a finite number of tanks the water can be placed into. As it links two physically distinguishable objects, $\rho$ is a continuous functions for reasoned discussed before.

As we combine the state of finite parts, we sum the distributions over particle states $\rho(\mathcal{s})=\rho_1(\mathcal{s})+\rho_2(\mathcal{s})$: the amount of material in the composite state is the sum of the amount of material of the parts. We can also increase or decrease the amount of material by a constant factor $\rho_1(\mathcal{s})=a\rho_2(\mathcal{s})$. This gives us the structure of a real vector space that is isomorphic to a subspace of continuous functions.

\begin{defn}\label{classical_vector space}
The state space $\mathcal{C}$ for a homogeneous classical (i.e. infinitesimally reducible) material is a vector space over $\mathbbm{R}$ isomorphic to a subspace of $C(\mathcal{S}) = \{\rho:\mathcal{S} \rightarrow \mathbbm{R} | \rho$ is continuous $\}$, where $\mathcal{S}$ is the state space of an infinitesimal amount of material.
\end{defn}

@@

Consider the set of transformations $T$ that increase or decrease the amount of material in the system by a constant. We claim $T$ is a field\footnote{Here field is intended in the abstract algebraic sense (a nonzero commutative division ring) which has no relationship to the field in the physics or differential geometry sense (a physical quantity/tensor with a value for each point in space).} isomorphic to $\mathbbm{R}$. Consider $\tau: \mathbbm{R} \rightarrow T$ the mapping between a number and the transformation that increases or decreases the amount of material by that number. This transformation exists: the system is infinitesimally decomposable and the amount can be changed continuously. Define on $T$ an addition $+: T \times T \rightarrow T$ and a multiplication $*: T \times T \rightarrow T$ such that $\tau(a) + \tau(b) = \tau(a+b)$ and $\tau(a) * \tau(b) = \tau(a*b)$, $a,b \in \mathbbm{R}$, so that the sum and product of the transformation is equal to the sum and product of their respective factors. $\tau$ is an isomorphism between $T$ and $\mathbbm{R}$ as fields.

We claim $\mathcal{C}$ is a vector space over $\mathbbm{R}$. The abelian group $\mathcal{C}$ can be extended with the operations defined by $T$, as each element $\tau \in T$ is a map $\tau : \mathcal{C} \rightarrow \mathcal{C}$. The map has the following properties: $(\tau_1 + \tau_2) \mathcal{c} = \tau_1 \mathcal{c} + \tau_2 \mathcal{c}$, increasing the amount of material by the sum of two constant is the same as combining the separate increases, and $\tau (\mathcal{c}_1 + \mathcal{c}_2) = \tau \mathcal{c}_1 + \tau \mathcal{c}_2$, increasing the amount of the total system is the same as the combination of the increased parts. $\mathcal{C}$ is a module over $T$, which is a field and isomorphic to $\mathbbm{R}$. $\mathcal{C}$ is (isomorphic to) a real vector space.

We claim $\mathcal{C}$ is isomorphic to a subspace of $C(\mathcal{S})$ as a vector space over $\mathbbm{R}$. $\forall \mathcal{c} \in \mathcal{C} \exists ! \rho_{\mathcal{c}}:\mathcal{S} \rightarrow \mathbbm{R}$ returning the amount of material for each state $\mathcal{s} \in \mathcal{S}$. $\rho \in C(\mathcal{S})$ by \ref{continuous_map}. Let $\varrho : \mathcal{C} \rightarrow C(\mathcal{S})$ such that $\varrho(\mathcal{c}) \mapsto \rho_\mathcal{c}$. 
As the system is reducible, two distinct composite state must represent different distributions. $\forall \mathcal{c_1}, \mathcal{c_2} \in \mathcal{C}, \mathcal{c_1} \neq \mathcal{c_2} \implies \varrho(\mathcal{c_1}) \neq \varrho(\mathcal{c_1})$. $\varrho$ is injective. $\varrho$ is a bijection between $\mathcal{C}$ and $\varrho(\mathcal{C})$. Let $\mathcal{c}=\mathcal{c}_1+\mathcal{c}_2$, then $\varrho(\mathcal{c})=\varrho(\mathcal{c}_1)+\varrho(\mathcal{c}_2)$ as the amount of material of the composition of two states is the sum of the individual amounts. Let $\mathcal{c}_1=\tau(a)\mathcal{c}_2$, then $\varrho(\mathcal{c}_1)=a \varrho(\mathcal{c}_2)$ as $\tau(a)$ increases the amount of material by the factor $a$. $\varrho$ is a homomorphism. $\mathcal{C}\cong \varrho(\mathcal{C}) \subseteq C(\mathcal{S})$ as vector spaces.
\end{justification}

@@

In the case where the topology of $\mathcal{S}$ is not discrete, the distribution $\rho_\mathcal{c}$ is a density which is not something we directly measure. What we do measure experimentally are finite amounts of material. For example, the amount of fluid in a particular volume within a particular range of momentum specified in some units (e.g. mass, kg, ...). This means that given a composite state $\mathcal{c}$ and a set $U \subseteq \mathcal{S}$ of particle states compatible with an outcome of a process, we must be able to tell the amount of material that we will find associated with that outcome. That is for each open set $U \subseteq \mathcal{S}$ there will be a functional $\Lambda_U : \mathcal{C} \rightarrow \mathbbm{R}$.

As we combine parts, the total amount of material will be the sum of their parts. As we increase or decrease by a constant factor the amount of material will increase or decrease by the same factor. That is: $\Lambda_U(a_1 \mathcal{c}_1 + a_2 \mathcal{c}_2) = a_1 \Lambda_U(\mathcal{c}_1) + a_2 \Lambda_U(\mathcal{c}_2)$. For a proper state, one that is not the difference of two states, we will expect a positive amount of material. So, if a density $\rho_\mathcal{c}(\mathcal{S})$ is positive everywhere, then the total amount is also positive. This makes $\Lambda_U$ a positive linear functional. As the amount of material we will measure is always finite, the functional applied to any composite state $\mathcal{c}$ over any set $U$ will be finite.

The ability to associate finite amount of material with sets of states allows us to define finite regions of $\mathcal{S}$ and compare them. Consider a region of position and momentum in phase space. If we are able to spread a finite amount of material into a uniform distribution, then we know we have a finite region. And the density will give us an indication of how big the region is: if we double the region, the density will half. In other words: because we are describing densities over $\mathcal{S}$, we are able to give a unique measure $\mu$ for the size of sets of $\mathcal{S}$, a metric $d\mu$ and our positive linear functionals become integrals: $\Lambda_U (\mathcal{c}) : \int_U \rho_{\mathcal{c}} d \mu$.

@@

\begin{defn}\label{integration}
	The state space $\mathcal{S}$ for the particles of a homogeneous classical material is endowed with a measure $\mu$. The state space $\mathcal{C}$ is isomorphic to the space of continuous functions that are integrable over any subset of $\mathcal{S}$. That is: $\mathcal{C} \cong C(\mathcal{S}) \cap  L^1_{ew}(\mathcal{S}, \mu)$ where $L^1_{ew}(\mathcal{S}, \mu) = \{ \rho : \mathcal{S} \rightarrow \mathbbm{R} \; | \; |\int_{U} \rho d\mu| < \infty \; \forall U \subseteq \mathcal{S}\}$.
\end{defn}

@@

On a state space with discrete topology and a single state variable we simply have $\Lambda_U (\mathcal{c}) : \sum \limits_{q=a}^b D(q)$. For one dimensional continuous states, we would simply expect $\Lambda_U (\mathcal{c}) : \int_a^b \rho (q) dq$. By a simple change of state variable, we see that this does not work: $\rho(q')= \rho(q) dq/dq'$. This would make $\rho$ not a function of just the state $\mathcal{s}$, but also of the state variables we are using. This is inconsistent with our previous definition. Moreover if the transformation is not differential, the density is not even well-defined. This tells us that $\mathcal{S}$ cannot be any manifold: it has to be one that allows coordinate independent densities.

To do that $\mathcal{S}$ has to be, first of all, a differentiable manifold. This way, at least, the densities is well defined for all state variables.

If $\rho$ is to remain unchanged when one state variable changes, then one or more other state variables have to change as well so that the Jacobian of the transformation is unitary. This means that we cannot change physical units to all the variables as we please: they are part of a unit system. For example, changing units of position will also change units of momentum. Let $\mathcal{Q}$ be the manifold charted by a the minimum set of state variables that define such system of units. An element of $\mathcal{S}$ will be a geometrical object in such space. As these are states for infinitesimal parts, it will be a local object in $\mathcal{Q}$. As we saw before, it cannot be a point as the density $\rho(q')= \rho(q) dq/dq'$ is not invariant. It cannot be a vector $v(q) e_q$, as the density $\rho(q', v')= \rho(q, v) (dq/dq')^2$ is not invariant. A covector $k(q) dq$ is fine as $\rho(q', k')= \rho(q, k) (dq/dq') (dq'/dq) = \rho(q, k)$. In other words: a covector is the simplest local geometrical object that allows coordinate independent densities. We can identify $\mathcal{S}$ as the cotangent bundle $T^*\mathcal{Q}$, the space of all possible covectors.

If $\mathcal{Q}$ is one dimensional, then the state is fully identified by a pair $(q, k)$. $d\mu$ is proportional to $dq \wedge dk$. We set $d\mu = \hbar dq \wedge dk$, where $\hbar$ is the proportionality constant that will depend on the unit chosen to count the possibilities of the pair $(q, k)$. The cardinality for states in a finite (i.e. compact) region $U \subseteq \mathcal{S}$ will be given by $\int_U \hbar dq \wedge dk$.

@@

The cardinality of possibilities for each d.o.f. (i.e. the wedge product within d.o.f) and the orthogonality of different d.o.f. (i.e. the scalar product across d.o.f.) must be the same regardless of the choice of state variables. We can express both requirements mathematically in a compact way. We first define conjugate momentum as $p_i=\hbar k_i$ and unified state variables as $\xi^a\equiv \{q^i, p_i\}$. Then we consider the canonical two-form $\omega$ given by the following components:
\begin{align*}
\omega_{ab} =  \left[
\begin{array}{cc}
0 & 1 \\
-1 & 0 \\
\end{array}
\right] \otimes I_n =
\left[
\begin{array}{cc}
0 & I_n \\
-I_n & 0 \\
\end{array}
\right] \\
\end{align*}
It returns the wedge product within a d.o.f. and the scalar product across. Requiring the invariance of this metric under state variables assures us the cardinalities of states and possibilities is well defined.

The state space $\mathcal{S}$ is the symplectic manifold $(T^*\mathcal{Q}, \omega)$. This is the simplest manifold that allows us to define coordinate invariant densities. The metric allows us to measure the cardinality of possibilities within and across independent d.o.f.

\begin{defn}\label{classical_phase_space}
The state space $\mathcal{S}$ for the particles of a homogeneous classical  system is a symplectic manifold form by a cotangent bundle $T^*\mathcal{Q}$ equipped with the canonical two-form $\omega$.
\end{defn}

@@

We should note that we are starting to be in a position similar to the one discussed in \ref{discrete_state_metric} for discrete states. We have both a topological space and a metric that allows to count states and possibilities. The only thing we started from was the idea of an infinitesimally reducible system. That lead to a state space $\mathcal{S}$ for the infinitesimal parts, on which we must be able to define state variable independent distributions. Which in turn gave us degrees of freedom made by pair of conjugate variables and the symplectic metric typical of classical phase space.

>>>>

Given that scientific reductionism (i.e. the idea of reducing physical systems and interactions to the sum of their constituent parts in order to make them easier to study) is at the heart of fundamental physics, we now explore how to characterize a system in terms of its components. That is, we want to study the relationship between the state space $\mathcal{C}$ of a composite system and the state space $\mathcal{S}$ of its parts. In general, this is quite a complicated thing to do, which requires intimate knowledge of the system at hand. So we simplify our problem and study a system made of infinitesimal homogeneous parts. What we'll find is that under the additional assumption that the system is infinitesimally reducible (i.e. its state is equivalent to the state of its parts) and that each part undergoes deterministic and reversible evolution, the motion is suitably described by the standard framework of classical Hamiltonian mechanics.

\subsection{Homogeneous decomposable systems}
The notion that a system is decomposable means the states are equipped with a rule of composition that allows one to write $\mathcal{c}=\mathcal{c}_1+\mathcal{c}_2$: the composite system is the sum of its parts. For example, the state of a ball is equal to the state of its left and right parts.

The notion that the system is homogeneous means that the states of the composite and of the parts are not unrelated: they are made of the same material.\footnote{Whether a system is homogeneous depends on context (e.g. air can be thought as homogeneous if the mixture of gases does not change in space or in time due to phase transitions or chemical processes) and one must check that such property is maintained by time evolution.} In fact, the state space $\mathcal{C}$ of all systems composed of such material will include the state of the system as well as the states of all its parts (i.e. $\mathcal{c}, \mathcal{c}_1, \mathcal{c}_2 \in \mathcal{C}$). Since combining any two systems made of a homogeneous material will always give us a system made of the same homogeneous material (e.g. combining elements made of water gives us another element made of water), the state space $\mathcal{C}$ is closed under composition.

@@

\begin{justification}
We claim $\mathcal{C}$ is an additive monoid. There exists a law of composition $+ : \mathcal{C} \times \mathcal{C} \rightarrow \mathcal{C}$ that takes two states and returns one that is the physical composition of the two. The domain and codomain match because the material is homogeneous. The law is commutative $\mathcal{c}_1 +\mathcal{c}_2 = \mathcal{c}_2+\mathcal{c}_1$ and associative $(\mathcal{c}_1 + \mathcal{c}_2) + \mathcal{c}_3 = \mathcal{c}_1 + (\mathcal{c}_2 + \mathcal{c}_3)$, as it does not matter in what order we physically compose the parts. There exists a unique zero element $\mathcal{c} + 0 = \mathcal{c}$ and it represents the physically empty state (i.e. no amount of material).

We claim $\mathcal{C}$ is an additive group. As we want to describe changes during the evolution (i.e. $\hat{\mathcal{c}} = \mathcal{c} +\delta \mathcal{c}$) we introduce an inverse $- : \mathcal{C} \rightarrow \mathcal{C}$ such that $\mathcal{c} + ( - \mathcal{c}) = 0 \forall \mathcal{c} \in \mathcal{C}$. Such inverse may introduce objects that do not properly represent a physical state, but a state change. A change of a physically distinguishable object is itself physically distinguishable therefore $\mathcal{C}$ is still a Hausdorff and second countable topological manifold as per \ref{topologically_distinguishable}. We'll still call $\mathcal{C}$ state space committing an abuse of terminology.
\end{justification}

@@

We will call a classical system one that is infinitesimally reducible. That is: it is made of arbitrarily small parts, which we call particles, and the state of the whole system coincides with the sum of the states of the parts. Given the state space $\mathcal{S}$ for the infinitesimal parts, a composite state $\mathcal{c}$ will tell us the amount of material for each possible particle state. That is, each state is fully identified by a function $\rho: \mathcal{S} \to \mathbbm{R}$ that returns the amount of material in the composite state $\mathcal{c}$ that is prepared in the particle state $\mathcal{s}$. For example, given a certain configuration of gas, we can tell the density of material that is at a specific point in space with a specific value of momentum.

The notion that the parts are infinitesimal requires the co-domain of $\rho$ to be continuous, as opposed to discrete. The state space $\mathcal{S}$, instead, can have a discrete topology. For example, for a system composed of different tanks connected by pipes, the state of the composite system could be the overall distribution of water among the tanks. The amount in each tank is continuous (as we assume the water to be infinitesimally divisible) yet there are only a finite number of tanks the water can be placed into. As it links two physically distinguishable objects, $\rho$ is a continuous function for reasons discussed before.

As we combine the states of different parts, we sum the distributions over particle states $\rho(\mathcal{s})=\rho_1(\mathcal{s})+\rho_2(\mathcal{s})$: the amount of material in the composite state is the sum of the amount of material of the parts. We can also increase or decrease the amount of material by a constant factor $\rho_1(\mathcal{s})=a\rho_2(\mathcal{s})$. This gives us the structure of a real vector space that is isomorphic to a subspace of continuous functions.

\begin{defn}\label{classical_vector space}
The state space $\mathcal{C}$ for a homogeneous classical (i.e. infinitesimally reducible) material is a vector space over $\mathbbm{R}$ isomorphic to a subspace of $C(\mathcal{S}) = \{\rho:\mathcal{S} \rightarrow \mathbbm{R} \; | \; \rho$ is continuous$\}$, where $\mathcal{S}$ is the state space of an infinitesimal amount of material.
\end{defn}

@@

Consider the set of transformations $T$ that increase or decrease the amount of material in the system by a constant factor. We claim $T$ is a field\footnote{Here field is intended in the abstract algebraic sense (a nonzero commutative division ring) which has no relationship to the field in the physics or differential geometry sense (a physical quantity/tensor with a value for each point in space).} isomorphic to $\mathbbm{R}$. Consider $\tau: \mathbbm{R} \rightarrow T$ the mapping between a number and the transformation that increases or decreases the amount of material by that number. This transformation exists: the system is infinitesimally decomposable and the amount can be changed continuously. Define on $T$ an addition $+: T \times T \rightarrow T$ and a multiplication $*: T \times T \rightarrow T$ such that $\tau(a) + \tau(b) = \tau(a+b)$ and $\tau(a) * \tau(b) = \tau(a*b)$, $a,b \in \mathbbm{R}$, so that the sum and product of the transformation is equal to the sum and product of their respective factors. $\tau$ is an isomorphism between $T$ and $\mathbbm{R}$ as fields.

We claim $\mathcal{C}$ is a vector space over $\mathbbm{R}$. The abelian group $\mathcal{C}$ can be extended with the operations defined by $T$, as each element $\tau \in T$ is a map $\tau : \mathcal{C} \rightarrow \mathcal{C}$. The map has the following properties: $(\tau_1 + \tau_2) \mathcal{c} = \tau_1 \mathcal{c} + \tau_2 \mathcal{c}$, increasing the amount of material by the sum of two constant factors is the same as combining the separate increases, and $\tau (\mathcal{c}_1 + \mathcal{c}_2) = \tau \mathcal{c}_1 + \tau \mathcal{c}_2$, increasing the amount of the total system is the same as the combination of the increased parts. $\mathcal{C}$ is a module over $T$, which is a field and isomorphic to $\mathbbm{R}$. $\mathcal{C}$ is (isomorphic to) a real vector space.

We claim $\mathcal{C}$ is isomorphic to a subspace of $C(\mathcal{S})$ as a vector space over $\mathbbm{R}$. $\forall \mathcal{c} \in \mathcal{C} \exists ! \rho_{\mathcal{c}}:\mathcal{S} \rightarrow \mathbbm{R}$ returning the amount of material for each state $\mathcal{s} \in \mathcal{S}$. $\rho \in C(\mathcal{S})$ by \ref{continuous_map}. Let $\varrho : \mathcal{C} \rightarrow C(\mathcal{S})$ such that $\varrho(\mathcal{c}) \mapsto \rho_\mathcal{c}$. 
As the system is reducible, two distinct composite states must represent different distributions. $\forall \mathcal{c_1}, \mathcal{c_2} \in \mathcal{C}, \mathcal{c_1} \neq \mathcal{c_2} \implies \varrho(\mathcal{c_1}) \neq \varrho(\mathcal{c_1})$. $\varrho$ is injective. $\varrho$ is a bijection between $\mathcal{C}$ and $\varrho(\mathcal{C})$. Let $\mathcal{c}=\mathcal{c}_1+\mathcal{c}_2$, then $\varrho(\mathcal{c})=\varrho(\mathcal{c}_1)+\varrho(\mathcal{c}_2)$ as the amount of material of the composition of two states is the sum of the individual amounts. Let $\mathcal{c}_1=\tau(a)\mathcal{c}_2$, then $\varrho(\mathcal{c}_1)=a \varrho(\mathcal{c}_2)$ as $\tau(a)$ increases the amount of material by the factor $a$. $\varrho$ is a homomorphism. $\mathcal{C}$ is isomorphic to $\varrho(\mathcal{C}) \subseteq C(\mathcal{S})$ as vector spaces.
\end{justification}

@@

In the case where the topology of $\mathcal{S}$ is not discrete, the distribution $\rho_\mathcal{c}$ is a density, which is not something we directly measure. What we do measure experimentally are finite amounts of material. For example, the amount of fluid in a particular volume within a particular range of momentum specified in some units (e.g. moles, kg, ...). This means that given a composite state $\mathcal{c}$ and a set $U \subseteq \mathcal{S}$ of particle states compatible with an outcome of a process, we must be able to tell the amount of material that we will find associated with that outcome. That is for each open set $U \subseteq \mathcal{S}$ there will be a functional $\Lambda_U : \mathcal{C} \rightarrow \mathbbm{R}$.

As we combine parts, the total amount of material will be the sum of the material of the parts. As we increase or decrease by a constant factor the total amount of material in a state, we will find the total amount of material of that state increased or decreased by that factor. That is: $\Lambda_U(a_1 \mathcal{c}_1 + a_2 \mathcal{c}_2) = a_1 \Lambda_U(\mathcal{c}_1) + a_2 \Lambda_U(\mathcal{c}_2)$. For a proper state, one that is not the difference of two states, we will expect a positive amount of material. So, if a density $\rho_\mathcal{c}$ is positive everywhere, then the total amount is also positive. This makes $\Lambda_U$ a positive linear functional. As the amount of material we will measure is always finite, the functional applied to any composite state $\mathcal{c}$ over any set $U$ will be finite.

The ability to associate finite amount of material with sets of states allows us to define finite regions of $\mathcal{S}$ and compare them. Consider a region of position and momentum in phase space. If we are able to spread a finite amount of material into a non-infinitesimal uniform distribution, then we know we have a finite region. And the density will give us an indication of how big the region is: if we double the region, the density will halve. In other words: because we are describing densities over $\mathcal{S}$, we are able to give a unique measure $\mu$ for the size of sets of $\mathcal{S}$, a metric $d\mu$ and our positive linear functionals become integrals: $\Lambda_U (\mathcal{c}) = \int_U \rho_{\mathcal{c}} d \mu$.

@@

\begin{defn}\label{integration}
	The state space $\mathcal{S}$ for the particles of a homogeneous classical material is endowed with a measure $\mu$. The state space $\mathcal{C}$ is isomorphic to the space of continuous functions that are integrable over any subset of $\mathcal{S}$. That is: $\mathcal{C} \cong C(\mathcal{S}) \cap  L^1_{ew}(\mathcal{S}, \mu)$ where $L^1_{ew}(\mathcal{S}, \mu) = \{ \rho : \mathcal{S} \rightarrow \mathbbm{R} \; | \;\; |\int_{U} \rho d\mu| < \infty \; \forall U \subseteq \mathcal{S}\}$.
\end{defn}

@@

On a state space with discrete topology and a single state variable we simply have $\Lambda_U (\mathcal{c}) = \sum \limits_{q=a}^b \rho(q)$. For states identified by a single continuous state variable we would expect $\Lambda_U (\mathcal{c}) = \int_a^b \rho (q) dq$. By a simple change of state variable, we see that this does not work in general: $\rho(q')= \rho(q) dq/dq'$. This would make $\rho$ a function not just of the state $\mathcal{s}$, but also of the state variables we are using. This is inconsistent with our previous definition. Moreover if the transformation is not differential, the density is not even well-defined. This tells us that $\mathcal{S}$ cannot be any manifold: it has to be one that allows state-variable-independent densities. Under arbitrary change of state variables, the density has to remain the same, which means that the Jacobian of the transformation must exist and must be unitary.

For the Jacobian to exist $\mathcal{S}$ has to be a differentiable manifold; that is the transformation between state variables is always differentiable. This way, at least, the density is well defined for any choice of state variables.

If $\rho$ is to remain unchanged and the Jacobian to be unitary we cannot simply change one state variable as we please. If we change one at least another has to change in some coordinated way such that the Jacobian of the total transformation is unitary. This means that we cannot change physical units of all the variables as we please: they are part of a unit system. For example, changing units of position will also change units of momentum. Let $\mathcal{Q}$ be the manifold charted by the minimum set of state variables that define such system of units. An element of $\mathcal{S}$ will be a geometrical object in such space. As these are states for infinitesimal parts, it will be a local object in $\mathcal{Q}$. As we saw before, it cannot be a point as the density $\rho(q')= \rho(q) dq/dq'$ is not invariant. It cannot be a vector $v(q) e_q$, as the density $\rho(q', v')= \rho(q, v) (dq/dq')^2$ is not invariant. A covector $k(q) dq$ is instead fine as $\rho(q', k')= \rho(q, k) (dq/dq') (dq'/dq) = \rho(q, k)$. In other words: a covector is the simplest local geometrical object that allows coordinate independent densities. We can identify $\mathcal{S}$ as the cotangent bundle $T^*\mathcal{Q}$, the space of all possible covectors.

If $\mathcal{Q}$ is one dimensional, then the state is fully identified by a pair $(q, k)$. These variables are independent, so the total number of possibilities in a small region of $\mathcal{S}$ is proportional to $dq \wedge dk$. That is if we double the range of $dq$ or $dk$ then the number of possibilities doubles. We set $d\mu = \hbar dq \wedge dk$, where $\hbar$ is the proportionality constant that will depend on the unit chosen to count the possibilities of the pair $(q, k)$. That is, a unitary range of possibilities for $q$ and $k$ will give us $\hbar$ possibilities.\footnote{The actual value and physical dimensions of $\hbar$ are determined by the system of units, and should not be taken to describe some intrinsic physical property. Only dimensionless relationships to other physical constants, such as the fine structure constant $\alpha = e^2/\hbar c$, can possess that trait. That is why one can choose "natural units" for which $\hbar=1$. In SI units the relationship $\hbar dk = dp = m du$ between kinetic and conjugate momentum, derived later, sets the relationship between units.} The cardinality for states in a finite (i.e. compact) region $U \subseteq \mathcal{S}$ will be given by $\int_U \hbar dq \wedge dk$.

@@

The cardinality of possibilities for each d.o.f. (i.e. the wedge product within a d.o.f) and the orthogonality of different d.o.f. (i.e. the scalar product across d.o.f.) must be the same regardless of the choice of state variables. We can express both requirements mathematically in a compact way. We first define conjugate momentum as $p_i=\hbar k_i$ and unified state variables as $\xi^a\equiv \{q^i, p_i\}$. Then we consider the canonical two-form $\omega$ given by the following components:
\begin{align*}
\omega_{ab} =  \left[
\begin{array}{cc}
0 & 1 \\
-1 & 0 \\
\end{array}
\right] \otimes I_n =
\left[
\begin{array}{cc}
0 & I_n \\
-I_n & 0 \\
\end{array}
\right] \\
\end{align*}
It returns the wedge product within a d.o.f. and the scalar product across. Requiring the invariance of this metric under state variable changes assures us the cardinality of states and possibilities is well defined.

This gives us insight on the physical meaning of the geometrical structure $T^*\mathcal{Q}$. The canonical one-form $\theta=k(q) dq$ represents the geometrical object we associate with each particle state, which is not a point in $\mathcal{Q}$. The canonical two-form $\omega=\Sigma \, dq^i \wedge dk_i$ is the metric that quantifies the possibilities described by two given state variables. Note, instead, that the relationship $\omega = - \hbar d\theta$, while mathematically true, has no clear physical meaning as we have not defined what the exterior derivative on a state actually represents.

We can capture the above discussion by stating that the state space $\mathcal{S}$ is the symplectic manifold $(T^*\mathcal{Q}, \omega)$. This is the simplest manifold that allows us to define state-variable-invariant densities. The metric allows us to measure the cardinality of possibilities within and across independent d.o.f.

\begin{defn}\label{classical_phase_space}
The state space $\mathcal{S}$ for the particles of a homogeneous classical  system is a symplectic manifold formed by a cotangent bundle $T^*\mathcal{Q}$ equipped with the canonical two-form $\omega$.
\end{defn}

@@

We should note that we are starting to be in a position similar to the one discussed in \ref{discrete_state_metric} for discrete states. We have both a topological space and a metric that allows us to count states and possibilities. We recovered the metric starting from the idea of an infinitesimally reducible system. That led to a state space $\mathcal{S}$ for the infinitesimal parts, on which we must be able to define state-variable-independent distributions, which in turn gave us degrees of freedom made by pairs of conjugate variables and the symplectic metric typical of classical phase space. In short: being able to measure the amount of material is what allows us to measure the cardinality of states.
